---
title: "package content"
author: "Zhuoer Dong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{package content}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = F,
  comment = ''
)

library(magrittr)
```

Still a draft, to be finished.

```{r}
library(hgnc)
```

after you read the `vignette('hgnc')` 

# 1

It achieves this by two modules: a generic interface which convert between "wide" map and "long" map.

`melt_map()`

`cast_map()`

`qualify_map()` make sure a map is "square". id-symbol has to fulfill this requirement

For both .., `soft` can include other columns than `ID` and `Ensembl`, which will just be ignored.

# 2


 to make things worse, the separator is not definite (it can be ` `, ` ã€`, ` \\\ `).

https://github.com/dongzhuoer/rGEO/blob/master/R/platform.R#L187-L194

# 3 

as_symbol

melt `hgnc_complete_set.txt.gz`

    for example, Entrez gene ID
   ```r
   library(magrittr)
   # the exact parameter should be character, but integer or numeric is also acceptable
   c(23141, 5925) %>% hgnc::as_symbol_from_entrez() # 
   ```

# other feature




to do

So, based on a pre-defined map between `id` and `symbol` (such as `entrez2symbol`), we can transform any user supplied GPL file to standard chip file ready for GSEA

