---
title: "Introduction to hgnc"
author: "Zhuoer Dong"
date: "`r Sys.Date()`"
output: prettydoc::html_pretty
vignette: >
  %\VignetteIndexEntry{Introduction to hgnc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = F,
  comment = ''
)

library(magrittr)
```

```{r}
library(hgnc)
```

# quick start



```{r make-probe2ensembl, include=FALSE}
probe2ensembl <- tibble::tribble(
  ~ID,           ~Ensembl,                                                 
  "200064_at",   "ENSG00000096384",                                        
  "200066_at",   "ENSG00000113141",                                        
  "200068_s_at", "ENSG00000127022",                                        
  "200069_at",   "ENSG00000075856",                                        
  "200071_at",   "ENSG00000119953",                                        
  "200076_s_at", "ENSG00000105700",                                        
  "200077_s_at", "ENSG00000104904",                                        
  "200078_s_at", "ENSG00000117410",                                        
  "200082_s_at", "ENSG00000171863 /// ENSG00000183405 /// ENSG00000213326",
  "200084_at",   "ENSG00000110696"
)
```

For impatient people, you can convert a probe-ensembl map like the following [^probe2ensembl]

```{r echo-probe2ensembl}
probe2ensembl
```

[^probe2ensembl]: how to create the `probe2ensembl` variable 

    You can copy the following code:

    ```{r make-probe2ensembl, eval=FALSE}
    ```

    Or create it from scratch (some packages are needed):
    
    ```r
    # read probe annotation from a GSE SOFT file
    soft_table <- system.file('extdata/GSE19161_family.soft.gz', package = 'rGEO') %>% 
        rGEO::parse_gse_soft(verbose = F) %>% {.$table}
    # subset part of the table for the purpose of demostration
    probe2ensembl <- soft_table %>% dplyr::select(1, 9) %>% dplyr::slice(41:50)
    ```

to a probe-symbol map ^[Ensembl IDs which don't have corresponding HUGO symbol are discarded.]

```{r}
probe2ensembl %>% melt_map('ID', 'Ensembl', ' /// ') %>%
    dplyr::mutate('symbol' = as_symbol_from_ensembl(Ensembl)) %>%
    cast_map('ID', 'symbol', ' /// ')
```

Other common ID like entrez gene ID, Unigene ID, RefSeq accession are also supported.


# innovation

## the problem

To master this package, you need to understand the problem it aims to solve. That is, to turn something like

```{r echo=FALSE}
tibble::tribble(  
	~probe, ~id,
    'probe1', 'id1 | id2',
    'probe2', 'id3'
) %>% knitr::kable()
```

to 

```{r echo=FALSE}
tibble::tribble(  
	~probe, ~symbol,
    'probe1', 'symbol1 | symbol2',
    'probe2', 'symbol3'
) %>% knitr::kable()
```

## solve a simpler one

To point out the key difficulty, let's contrast it with a simpler one --- to turn something like 

```{r echo=FALSE}
tibble::tribble(  
	~probe, ~id,
    'probe4', 'id4',
    'probe5', 'id5',
    'probe6', 'id6'
) %>% knitr::kable()
```

to 

```{r echo=FALSE}
tibble::tribble(  
	~probe, ~symbol,
    'probe4', 'symbol4',
    'probe5', 'symbol5',
    'probe6', 'symbol6'
) %>% knitr::kable()
```

That's quite easy, you just need a id-symbol map

```{r echo=FALSE}
tibble::tribble(  
	~id, ~symbol,
    'id4', 'symbol4',
    'id5', 'symbol5',
    'id6', 'symbol6'
) %>% knitr::kable(caption = 'id2symbol')
```

and use the following code ^[I deliberately shuffle the rows of `id2symbol` to show that `id2symbol` just need to provide the correct relationship between `id` and `symbol`, i.e, it doesn't necessarily maintain the same order as `probe2id`.]

```{r}
probe2id <- tibble::tribble(  
	~probe, ~id,
    'probe4', 'id4',
    'probe5', 'id5',
    'probe6', 'id6'
)

id2symbol <- tibble::tribble(  
	~id, ~symbol,
    'id5', 'symbol5',
    'id6', 'symbol6',
    'id4', 'symbol4'
) 

probe2symbpl <- probe2id %>% 
    dplyr::transmute(probe, symbol = id2symbol$symbol[match(id, id2symbol$id)]) %T>% print
```

In the above code, we map `probe` to `symbol` by three steps [^probe4-symbol4].

[^probe4-symbol4]: In general:

    a. `dplyr::transmute` preserves `probe2id$probe` - `probe2id$id` relationship by _position_
    b. `match()` finds `probe2id$id` - `id2symbol$id` relationship by _value_
    c. `[]` finds `id2symbol$id` - `id2symbol$symbol` relationship by _position_

    Let's us experience this by an example of `'probe4'` to `'symbol4'`:

    a. 1st element of `probe2id$probe` -> 1st element of `probe2id$id`:  
       `'probe4'` is the 1st element of `probe2id$probe`, so we look for the 1st element of `probe2id$id`, `'id4'`.
  
    b. `'id4'` in `probe2id$id` -> `'id4'` in `id2symbol$id`:  
       `'id4'` is the 1st element of `probe2id$id`, then we look for the 1st element of `match()` (which gives the position of `probe2id$id` in `id2symbol$id` --- ``c(`r match(probe2id$id, id2symbol$id)`)`` in this case). We get `3`, so we look for the 3rd element of `id2symbol$id`, the exact value of `'id4'`.
  
    c. 3rd element of `id2symbol$id` -> 3rd element of `id2symbol$symbol`:  
       finally, `'id4'` is the 3rd element of ``id2symbol$id`, thus we look for the 3rd element of `id2symbol$symbol`, `'symbol4'`.


## key difficulty 

Back the original problem, you can find that its fairly easy to "replace" `'id4'` with `'symbol4'`, `'id5'` with `'symbol5'`, etc (thanks to `match()`). But how can you "replace" the `'id1'` and `'id2'` inside `'id1 | id2'`? 

That is what I meet when I want to address a [biological question](https://github.com/dongzhuoer/rGEO), as you can see in the 9th line of `probe2ensembl`.

```{r}
probe2ensembl %>% dplyr::slice(9)
```

If you think it's a piece of cake, you may have some misunderstanding [^not-easy]. 

[^not-easy]: 
    - Computer is very foolish, it can't convert `'id1 | id2'` to `'symbol1 | symbol2'` as you can easily achieve even without thinking. In programming, the only way is to search `'id1'` in `'id1 | id2'` and replace with `'symbol'` if it find one, then search `'id2'`, `'id3'`, etc. This will cause a severe poor performance.

    - As for replacing all `'id'` with `'symbol'`, I use `id2symbol` just to simplify the problem, the id-symbol map in the real world is usually something like:

      ```{r}
      hgnc::ensembl2symbol %>% dplyr::sample_n(10)
      ```

## solution

Inspired by [**reshape2**](https://github.com/hadley/reshape), I choose to "melt" 

```{r}
probe2id_wide <- tibble::tribble(  
	~probe, ~id,
    'probe1', 'id1 | id2',
    'probe2', 'id3'
)
```

to 

```{r}
probe2id_long <- probe2id_wide %>% melt_map('probe', 'id', ' \\| ') %T>% print
```

I call the former "wide" map and the latter "long" map.

Then use the code in [simpler problem](#solve-a-simpler-one) to map id to symbol

```{r}
id2symbol <- tibble::tribble(  
	~id, ~symbol,
    'id1', 'symbol1',
    'id2', 'symbol2',
    'id3', 'symbol3'
) 

probe2symbol_long <- probe2id_long %>% 
    dplyr::transmute(probe, symbol = id2symbol$symbol[match(id, id2symbol$id)]) %T>% print
```

Finally "cast" to 

```{r}
probe2symbol_wide <- probe2symbol_long %>% cast_map('probe', 'symbol', ' /// ') %T>% print
```

In short, A-B wide map -> A-B long map -> A-C long map -> A-C wide map.

